Brief Announcement: An Improved Lower Bound for Dynamic Graph Partitioning

### 2020-01-28 paper/paper.tex
PPL and its proof

@@

Formally, the online \emph{Balanced RePartitioning} problem (BRP) is defined as
follows. There is a set of $n$ nodes, initially distributed arbitrarily
across $\ell$~clusters, each of size~$k$. We call two nodes~$u,v\in V$
\emph{collocated} if they are in the same cluster.

@@

\begin{proof}
We construct an instance of the problem with $\ell$ clusters 
$\set{ S_1, S_2,\dots , S_{\ell}}, |S_i|  = k$.
Let $I(C)$ denote the cluster where nodes in a component $C$ are located initially.
Consider any online algorithm ALG.
We construct the input sequence of requests for ALG as follows.
First,
we issue $k-2$ (internal) requests so that ALG form a component of $k-1$
nodes on clusters $S_1$.
Note that this does not cost ALG.
Let $x_0$  be the only single nodes left on $S_1$ and  $y_0 \in S_2$ be any single node on $S_2$.
Next,
we issue the request between $x_0$ and $y_0$.
Since this is an external request,
ALG joins them into one component and collocates them in some cluster other than $S_1$.
For this,
ALG moves to a new configuration that replaces $x_0$ and $y_0$ with two other single nodes $x_1$ and $y_1$ respectively.
Next,
we issue a request between $x_1$ and the largest component $C$ s.t.~$I(C) = I(x_1)$.
ALG must collocate $x_1$ and $C$ on some cluster other than $S_1$ and
consequently replaces $x_1$.
We repeat issuing requests between the single node $x_i$ on $S_1$ and the largest component $C'$ s.t.~$I(C')=I(x_i)$,
 until there are only two single nodes left that  originate from the same cluster.
I.e.,
two single nodes $x^*, y^*,I(x^*) = I(y^*)$
that constitute the last pair of such nodes.
At this point there are at most $\ell+1$ single nodes left
since there must be at least two nodes with the same initial cluster in any subset of $\ell+1$
nodes.
Given this sequence of requests,
the optimal strategy is to migrate $\set{x_0,y_0}$ to the cluster $I(x^*)$ by
 swapping $x_0$ and $y_0$ with $x^*$ and $y^*$ respectively.
Hence,
OPT pays $4$ for node migrations and
ALG pays at least one for each node in the sequence $X := x_0, x_1,\dots$.
We exclude at most $(k-1) + ( \ell+1)$ nodes out of $k.\ell$ nodes,
therefore $|X| \geq k.\ell - k - \ell \in \Omega(k.\ell)$.
\end{proof}

@@

Let $P_I = I_1, \dots, I_{\ell}$ denote the initial partition with which both ON and OPT begin and
$P_F = F_1, \dots, F_{\ell}$ the final partition.
\begin{definition}	\label{def:dist}
    The \emph{distance} of a partition $P = C_1, \dots, C_{\ell}$ is the number of nodes in $P$ that do not reside in their initial cluster.
    That is,
    $dist(P, P_I) = \sum_{j=1}^{\ell} | C_j \setminus I_j |$. 
\end{definition}
\maciek{We should figure out a better name than "distance". We usually apply two arguments to distance. Examples of names that can be applied to only one configuration as a parameter: radius, span, stretch, spread, width. Also: difference, disparity, distinction, alternation}

In other words,
at least $dist(P, P_I)/2$ node swaps are required in order to reach the partition $P$ from $P_I$.
Therefore,
$OPT \geq \Delta:= dist(P_F, P_I) $.
We devise an online algorithm that mimics OPT by minimizing the distance to the initial partition $P_I$ with each re-partitioning.
As a result,
ON never ends up in a partition that is further away from $P_I$ than $\Delta$.
This invariant ensures that ON does not pay too much while reaching $P_F$.

\begin{algorithm}
    \renewcommand{\algorithmicrequire}{\textbf{Input:}}
    \renewcommand{\algorithmicensure}{\textbf{Output:}}
    \begin{algorithmic}[1]
        \Require 
        $k, \ell$,
        initial partition $P_I$,
        sequence of  edges $\sigma_1, \dots, \sigma_N$ 
        \Ensure Final partition $P_F$ 
        \State for each node $v$ create a singleton component $C_v$ and add it to $\mathcal{C}$ \label{line:initcomponents}
        \State on communication request $\sigma_t=\{u,v\}$:
        \State Let $C_1 \ni u$ and $C_2 \ni v$ be the container components
        \If{$C_1 \neq C_2$}
        \State unite the two components into a single component $C'$ and
        $\mathcal{C} = (\mathcal{C}\setminus\set{C_1, C_2}) \cup ~\set{C'}$ \label{line:mergecomponents}
        \If{cluster$(C_1) \neq$ cluster$(C_2)$}		
        \State \textit{re-partition($k, \ell, P_I, \mathcal{C}$)} \label{line:rebalance} 
        \EndIf
        \EndIf
    \end{algorithmic}
    \caption{Perfect Partition Learner}
    \label{alg:ppl}
      \end{algorithm}

@@

\mahmoud{address the \textit{re-partition()}}

\begin{property} \label{prop:dist<OPT}
    Let $P$ be any partition chosen by Algorithm \ref{alg:ppl} at Line $\ref{line:rebalance}$.
    Then, $dist(P,P_I) \leq \Delta$.
\end{property}

\begin{lemma}	\label{lemma:rebalancecost}
    The cost of re-partitioning at Line \ref{line:rebalance} is at most $2.OPT$.
\end{lemma}
\begin{proof}
    Consider the re-partitioning that transforms $P_{t-1}$ to $P_t$ upon the request $\sigma_t$.
    Let $M \subset V$ denote the set of nodes that are migrated during this process.
    We have $M = M^+ \cup M^- \cup M^\circ$,
    where $M^-$ ($M^+$) denotes the subset of nodes that
    enter (leave) their original cluster during the re-partitioning,
    and $M^\circ$ denotes the set of remaining nodes in $M$.
    Since $|M^- \cup M^\circ|$ nodes are not in their original cluster before the re-partitioning,
    by Definition \ref{def:dist},
    the distance before the re-partitioning is $dist(P_{t-1}) \geq | M^- \cup M^\circ |$.
    Analogously,
     the distance afterwards is $dist(P_{t}) \geq | M^+ \cup M^\circ |$.
    Thus,
    $|M| \leq dist(P_{t-1}) + dist(P_{t})$.
    By Property \ref{prop:dist<OPT},
    $dist(P_{t-1}) , dist(P_{t}) \leq \Delta \leq OPT$
    and thereby we have	
    $|M| \leq 2.OPT$.
\end{proof}

\begin{theorem}	\label{thm:upperbound}
    Algorithm \ref{alg:ppl} reaches the final partition while being $(2.k.\ell)$-competitive.
\end{theorem}
\begin{proof}
    The algorithm eventually reaches the final partition since it
     evaluates all possible $\ell$-way partitions of components on each external request,
    including the request that completes revealing of all ground truth components.
    There are at most $(k-1).\ell < k.\ell $ calls to \emph{re-partition()} and by Lemma \ref{lemma:rebalancecost} each costs at most $2.OPT$.
    The total cost is therefore at most $2.OPT.k.\ell$ which implies the competitive ratio.
      \end{proof}

@@

   Let \emph{Arb-partition} be the algorithm that moves to an arbitrary perfect partition
on each inter-cluster request.
\begin{theorem}
	Algorithm \emph{Arb-partition} is $O(\ell)$-competitive when $k=3$.
\end{theorem}

>>>>

\mathchardef\mhyphen="2D
\newcommand{\OPT}{\mathit{OPT}}

@@

Formally, the online \emph{Balanced Re-Partitioning} problem (BRP) is defined as
follows. There is a set of $n$ nodes, initially distributed arbitrarily
across $\ell$~clusters, each of size~$k$. We call two nodes~$u,v\in V$
\emph{collocated} if they are in the same cluster.

@@

\begin{proof}
We construct an instance of the problem with $\ell$ clusters 
$\set{ S_1, S_2,\dots , S_{\ell}}, |S_i|  = k$.
Let $I(C)$ denote the cluster where nodes in a component $C$ are located initially.
Consider any online algorithm ALG.
We construct the input sequence of requests for ALG as follows.
First,
we issue $k-2$ (internal) requests so that ALG form a component of $k-1$
nodes on clusters $S_1$.
Note that this does not cost ALG.
Let $x_0$  be the only single nodes left on $S_1$ and  $y_0 \in S_2$ be any single node on $S_2$.
Next,
we issue the request between $x_0$ and $y_0$.
Since this is an external request,
ALG joins them into one component and collocates them in some cluster other than $S_1$.
For this,
ALG moves to a new configuration that replaces $x_0$ and $y_0$ with two other single nodes $x_1$ and $y_1$ respectively.
Next,
we issue a request between $x_1$ and the largest component $C$ s.t.~$I(C) = I(x_1)$.
ALG must collocate $x_1$ and $C$ on some cluster other than $S_1$ and
consequently replaces $x_1$.
We repeat issuing requests between the single node $x_i$ on $S_1$ and the largest component $C'$ s.t.~$I(C')=I(x_i)$,
 until there are only two single nodes left that  originate from the same cluster.
I.e.,
two single nodes $x^*, y^*,I(x^*) = I(y^*)$
that constitute the last pair of such nodes.
At this point there are at most $\ell+1$ single nodes left
since there must be at least two nodes with the same initial cluster in any subset of $\ell+1$
nodes.
Given this sequence of requests,
the optimal strategy is to migrate $\set{x_0,y_0}$ to the cluster $I(x^*)$ by
 swapping $x_0$ and $y_0$ with $x^*$ and $y^*$ respectively.
Hence,
OPT pays $4$ for node migrations and
ALG pays at least one for each node in the sequence $X := x_0, x_1,\dots$.
We exclude at most $(k-1) + ( \ell+1)$ nodes out of $k.\ell$ nodes,
therefore $|X| \geq k.\ell - k - \ell \in \Omega(k\cdot\ell)$.
\end{proof}

@@

We assume OPT begins with the initial partition
$P_I = I_1, \dots, I_{\ell}$ and moves to the final partition
$P_F = F_1, \dots, F_{\ell}$.
 The \emph{distance} of a partition $P = C_1, \dots, C_{\ell}$ from the initial partition
 is the number of nodes in $P$ that do not reside in their initial cluster.
    That is,
    $\mathit{dist}(P, P_I) = \sum_{j=1}^{\ell} | C_j \setminus I_j |$. 
In other words,
at least $\mathit{dist}(P, P_I)/2$ node swaps are required in order to reach the partition $P$ from $P_I$.
Therefore,
$\OPT \geq \Delta:= dist(P_F, P_I) $.
The online algorithm PPL mimics OPT by minimizing the distance to the initial partition $P_I$ with each re-partitioning.
As a result,
ON never ends up with a partition that is more than $\Delta$ away from $P_I$.
This invariant ensures that ON does not pay too much while reaching $P_F$.

\maciek{We should figure out a better name than "distance". We usually apply two arguments to distance. Examples of names that can be applied to only one configuration as a parameter: radius, span, stretch, spread, width. Also: difference, disparity, distinction, alternation}

\begin{algorithm}
    \renewcommand{\algorithmicrequire}{\textbf{Input:}}
    \renewcommand{\algorithmicensure}{\textbf{Output:}}
    \begin{algorithmic}[1]
        \Require 
        $k, \ell$,
        initial partition $P_I$,
        sequence of  requests $\sigma_1, \dots, \sigma_N$ 
        \Ensure A final partition $P_F$ 
        \State For each node $v$ create a singleton component $C_v$ and add it to $\mathcal{C}$
        \State{$P_0 := P_I$}
         \label{line:initcomponents}
        \For {each  request $\sigma_t=\{u,v\}, 1 \leq t \leq N$}
        \State Let $C_1 \ni u$ and $C_2 \ni v$ be the container components
        \If{$C_1 \neq C_2$}
        \State Unite the two components into a single component $C'$ and
        $\mathcal{C} = (\mathcal{C}\setminus\set{C_1, C_2}) \cup ~\set{C'}$ \label{line:mergecomponents}
%        \If{$\mathit{cluster}(C_1, P_{t-1}) \neq \mathit{cluster}(C_2, P_{t-1})$}
        \If{$C_1$ and $C_2$ are not in the same cluster w.r.t.~$P_{t-1}$}
        \State $P_{t} = \mathit{re\mhyphen partition}(P_{t-1}, P_I, \mathcal{C}$) \label{line:rebalance} 
        \Comment{move to $P$ closest to $P_I$}
        \EndIf
        \EndIf
        \EndFor
    \end{algorithmic}
    \caption{Perfect Partition Learner (PPL)}
    \label{alg:ppl}
      \end{algorithm}

@@

      We note that the $\mathit{re\mhyphen partition}$ at Line \ref{line:rebalance} replaces the current partition $P$ with a (perfect) partition closest to $P_I$.
Hence it never moves to a partition beyond distance $\Delta$.      
\begin{property} \label{prop:dist<OPT}
    Let $P$ be any partition chosen by Algorithm \ref{alg:ppl} at Line $\ref{line:rebalance}$.
    Then, $\mathit{dist}(P,P_I) \leq \Delta$.
\end{property}

\begin{lemma}	\label{lemma:rebalancecost}
    The cost of re-partitioning at Line \ref{line:rebalance} is at most $2\cdot\OPT$.
\end{lemma}
\begin{proof}
    Consider the re-partitioning that transforms $P_{t-1}$ to $P_t$ upon the request $\sigma_t$.
    Let $M \subset V$ denote the set of nodes that migrate during this process.
	Let $M^-$ and $M^+$ denote the subset of nodes that (respectively)
    enter or leave their original cluster during the re-partitioning.    
    Then,
    $M = M^+ \cup M^-$.
    Since $|M^-|$ nodes are not in their original cluster before the re-partitioning (i.e., in $P_{t-1}$),
    the distance before the re-partitioning is $\mathit{dist}(P_{t-1},P_I) \geq | M^-|$.
    Analogously,
     the distance afterwards is $\mathit{dist}(P_{t},P_I) \geq | M^+|$.
    Thus,
    $|M| \leq \mathit{dist}(P_{t-1},P_I) + \mathit{dist}(P_{t},P_I)$.
    By Property \ref{prop:dist<OPT},
    $\mathit{dist}(P_{t-1},P_I) , \mathit{dist}(P_{t},P_I) \leq \Delta \leq \OPT$
    and thereby we have	
    $|M| \leq 2\cdot\OPT$.
\end{proof}

\begin{theorem}	\label{thm:upperbound}
    Algorithm \ref{alg:ppl} reaches the final partition $P_F$ and it is $(2\cdot k\cdot\ell)$-competitive.
\end{theorem}
\begin{proof}
      On each inter-cluster request,
     the algorithm enumerates all $\ell$-way partitions of components
     that are in the same (closest) distance of $P_I$.
     That is, 
     once it reaches $\Delta = \mathit{dist} (P_{t}, P_I)$,
     it does not move to a partition
     $P_{t'}, t'>t,\mathit{dist} (P_{t'}, P_I) > \Delta$
     before it enumerates all partitions at distance $\Delta$.
     Therefore,
     PPL eventually reaches $\Delta=\OPT$ and the partition $P_F$.
%    including the request that completes revealing of all components that are collocated in $P_F$.
    There are at most $(k-1)\cdot\ell < k\cdot\ell $ calls   to $\mathit{re\mhyphen partition}$
     (i.e., the number of internal edges in $P_F$).
    By Lemma \ref{lemma:rebalancecost},
    each costs at most $2\cdot\OPT$.
    The total cost is therefore at most $2\cdot\OPT\cdot k\cdot\ell$ which implies the competitive ratio.
    \mahmoud{This is the cost of moving and the cost of remote comm. is not counted.
    	So it is 4-competitive (?)}
 \end{proof}

@@

   Let \emph{Arb-partition} be the algorithm that moves to an arbitrary perfect partition
on each inter-cluster (i.e., external) request.
\begin{theorem}
	Algorithm \emph{Arb-partition} is $O(\ell)$-competitive when $k=3$.
\end{theorem}
